---
title: "MI_May 2026"
author: "Ratna"
format: html
editor: visual
---

```{r}
library(mice) #for MI (creating the imputed datasets, method vector, prediction matrix, convergence plots)
library(miceadds) #provides 2l.norm
library(dplyr) #useful for data wrangling - mutate, select, filter, group_by
library(tidyr) #for reshaping data into LONG format - complete() comes from here
library(lme4) #to run the lmer regression models
library(broom.mixed) #needed for pool() with lmer output
library(ggplot2) #plots, visualizations
library(datawizard) #for standardization
library(effectsize)
library(purrr) #for imap
library(knitr)
library(easystats)
library(stringr)
library(interactions)
library(marginaleffects)
library(jtools)
```

#Load the input file

```{r}
df <- read.csv("Input for MI.csv")
```

#Calculate (a)mean for the predictor variables and add these new columns to df

```{r}

df <-  df %>%
        group_by(connectionId) %>%
        mutate(
        #Mean of posemo and negemo
        posemo_mean = mean(posemo, na.rm = TRUE),
        negemo_mean = mean(negemo, na.rm = TRUE),
        
        #mean of positive and negative diversity
        posdiv_mean = mean(positive_diversity_Shannons_lem, na.rm = TRUE),
        negdiv_mean = mean(negative_diversity_Shannons_lem, na.rm = TRUE),
        
        #mean of positive and negative granularity
        posgran_mean = mean(positive_granularity_lemm, na.rm = TRUE),
        neggran_mean = mean(negative_granularity_lemm, na.rm = TRUE),
        
        #mean of valence
        valence_mean = mean(valence, na.rm = TRUE)
        ) %>%
        ungroup()
```

#Input for MI

```{r}
mice_input_df <- df %>%
  filter(!is.na(connectionId)) %>%    # ← remove 4 NA rows
  filter(!is.na(valence)) %>% #remove rows without valence
  select(connectionId,
         alias,
         
         #raw variables
         valence,
         posemo,
         negemo,
         positive_granularity_lemm,
         negative_granularity_lemm,
         positive_diversity_Shannons_lem,
         negative_diversity_Shannons_lem,
         
         #person level means for predictors
         valence_mean,
         posemo_mean,
         negemo_mean,
         posgran_mean,
         neggran_mean,
         posdiv_mean,
         negdiv_mean
        ) %>%
    mutate(
          posemo_negemo = posemo * negemo,
          posdiv_negdiv = positive_diversity_Shannons_lem * negative_diversity_Shannons_lem,
          posgran_neggran =  positive_granularity_lemm * negative_granularity_lemm
    )
```

```{r}
#Create method vector
meth <- rep("", ncol(mice_input_df)) #creates an empty vector of the same length as the number of columns in the dataset
names(meth) <- colnames(mice_input_df) #assigns the column names of the dataset as the names of the meth vector
#Predict positive and negative granularity values using 2l.norm (two-level heteroskedastic model)
meth[c("positive_granularity_lemm", "negative_granularity_lemm")] <- "2l.norm" #for all the columns in the dataset, we will use the 2l.norm method for imputation

#Passively impute all interaction terms with missing values
#This is the only interaction whose components have missing values
meth["posgran_neggran"] <- "~I(positive_granularity_lemm * negative_granularity_lemm)"

#Set valence to "" so it is only used as predictor/filter, not imputed
meth[c("valence")] <- ""  #already "" by default, but made explicit for clarity
```

```{r}
#Create Prediction Matrix
pred <- make.predictorMatrix(mice_input_df)

#Set connectionId as cluster variable for multilevel imputation
pred[, "connectionId"] <- -2 
    #setting -2 in the predictor matrix identifies it as the cluster so it can calculate the "between-person" and "within-person" variance 

#Don't predict granularity values from their interaction 
# To prevent circular prediction : posgran_neggran is derived FROM these,
# so it cannot be used to predict them
pred["positive_granularity_lemm", "posgran_neggran"] <- 0
pred["negative_granularity_lemm", "posgran_neggran"] <- 0

#Don't predict gran terms as well as their interaction from "valence"
pred["positive_granularity_lemm", "valence"] <- 0
pred["negative_granularity_lemm", "valence"] <- 0
pred["posgran_neggran", "valence"] <- 0

#Don't predict granularity from itself
pred["positive_granularity_lemm", c("positive_granularity_lemm", "negative_granularity_lemm")] <- 0
pred["negative_granularity_lemm", c("positive_granularity_lemm", "negative_granularity_lemm")] <- 0

#Don't predict the gran interaction variable from either of them
pred["posgran_neggran", c("positive_granularity_lemm", "negative_granularity_lemm")] <- 0
```

```{r}
#Create an empty post-processing matrix
#Postprocessing ensures that the imputed granularity values stay within valid range
post <- make.post(mice_input_df)
```

\`

```{r}
#Create m imputed datasets 
imputed_data <- mice(mice_input_df,
    m = 100, #meaning there will be m unique guesses/imputed datasets which need to               be pooled later to one value across all m using Rubin's rules
    maxit = 50, #(with 50 iterations per dataset)
    predictorMatrix = pred,
    method = meth,
    post = post,
    printFlag = FALSE,
    seed = 2023
)
```



```{r}
#Check for convergence
plot(imputed_data)
```


#Calculate person mean centered values grouping by each m and connectionId
```{r}
#1) Create long format with pmc values
long_pmc <- mice::complete(imputed_data, action = "long", include = TRUE) %>%
                       group_by(.imp, connectionId)    %>% 
                       mutate(
                            # Recalculate person means POST imputation
                            posgran_mean_updated = mean(positive_granularity_lemm, na.rm = TRUE),
                            neggran_mean_updated = mean(negative_granularity_lemm, na.rm = TRUE),
                         
                         
                            #PMC after MI
                            #Person-mean-centered granularity 
                            posgran_pmc = positive_granularity_lemm - posgran_mean_updated,
                            neggran_pmc = negative_granularity_lemm - neggran_mean_updated,
                            
                            #Person-mean-centering for the predictor variable - diversity
                            posdiv_pmc = positive_diversity_Shannons_lem - posdiv_mean,
                            negdiv_pmc = negative_diversity_Shannons_lem - negdiv_mean,
                            
                            #Person-mean-centering for posemo, negemo
                            posemo_pmc = posemo - posemo_mean,
                            negemo_pmc = negemo - negemo_mean,
                            
                            #Interaction terms using PMC
                            posemo_negemo_pmc = posemo_pmc * negemo_pmc,
                            posdiv_negdiv_pmc = posdiv_pmc * negdiv_pmc,
                            posgran_neggran_pmc =  posgran_pmc * neggran_pmc
                       ) %>%
                       ungroup()

#2) Convert to mids
imputed_data_pmc <- mice::as.mids(long_pmc)
```


#Replace the values with NaNs for texts with 0 words
```{r}
#Add original_row_id to df
df <- df %>% dplyr::mutate(original_row_id = row_number())

#Extract the imputed data in long format
imputed_datasets <- mice::complete(imputed_data_pmc, action = "long", include = TRUE) %>%
    dplyr::left_join(
    df %>%
      dplyr::mutate(.id = row_number()) %>%
      dplyr::select(.id,
                    positive_all_words_count_lemm,
                    negative_all_words_count_lemm),
    by = ".id"
  ) %>%
  dplyr::mutate(
    posgran_pmc = dplyr::if_else(
      positive_all_words_count_lemm == 0, NA_real_, posgran_pmc),
    neggran_pmc = dplyr::if_else(
      negative_all_words_count_lemm == 0, NA_real_, neggran_pmc)
  ) %>%
  dplyr::select(-positive_all_words_count_lemm,   # ← drop after use
                -negative_all_words_count_lemm)    # ← no longer needed

#Convert back to mids
imputed_data_clean <- mice::as.mids(imputed_datasets)
```


##Standardizing the filtered imputed data
```{r}
# Step 1: Extract long format
long_data <- mice::complete(imputed_data_clean, action = "long", include = TRUE)

# Step 2: Standardize within each .imp separately
long_data_std <- long_data %>%
  group_by(.imp) %>%
  group_modify(~ datawizard::standardize(.x, select = c("posgran_pmc", 
                                                         "neggran_pmc",
                                                         "posemo_pmc", 
                                                         "negemo_pmc",
                                                         "posdiv_pmc", 
                                                         "negdiv_pmc"))) %>%
  ungroup()

# Step 3: Convert back to mids
imputed_data_std <- mice::as.mids(long_data_std)
```


#Null Model
```{r}
#Fit Null model
nulmodel <- with(imputed_data_std, lme4::lmer(valence ~ 1 + (1|connectionId)))

#Extract ICC from each imputed model and average
iccs <- sapply(nulmodel$analyses, function(m) {
 performance::icc(m)$ICC_adjusted
})

mean(iccs) #pooled ICC
sd(iccs) #Variability across imputations
```


#(CONTROLLING FOR AFFECTIVE TONE)
```{r}
#Fit the model - Text Level
mlm_v_s3 <- with(imputed_data_std, lme4::lmer(valence ~ posgran_pmc + #Individual Variables
                                                        neggran_pmc +
                                                        posemo_pmc + 
                                                        negemo_pmc +
                                                        posdiv_pmc +
                                                        negdiv_pmc +
                                                       
                                                        #Interaction Terms (Calculated post MI)
                                                        posemo_pmc * negemo_pmc +
                                                        posdiv_pmc * negdiv_pmc +
                                                        posgran_pmc * neggran_pmc +
                                              
                                                        #Random effect (texts nested within person)
                                                        + (1 | connectionId)))


#Pool them
s3_pool <- mice::pool(mlm_v_s3)

#View the final results with confidence intervals
s3_pool_results <- summary(s3_pool, conf.int = TRUE)
s3_pool_results
```



#MODEL(NOT CONTROLLING FOR AFFECTIVE TONE)
```{r}
main_1 <- with(imputed_data_std, lme4::lmer(valence ~
                                            
                                            #Main effects of interest
                                            posgran_pmc +
                                            neggran_pmc +
                                            posdiv_pmc +
                                            negdiv_pmc +
                                              
                                            #Interactions of interest
                                            posgran_pmc * neggran_pmc +
                                            posdiv_pmc * negdiv_pmc +
                                              
                                            #Random effect
                                            (1 | connectionId)))

main_1_pool <- mice::pool(main_1)
main_1_pool_results <- summary(main_1_pool, conf.int = TRUE)
main_1_pool_results
```


#Beta and p values for all interactions WITHOUT and WITH AFFECTIVE TONE
```{r}
# WITHOUT Affective Tone - interactions only
print("WITHOUT Affective Tone - Interactions:")
main_1_pool_results[main_1_pool_results$term %in% 
                    c("posgran_pmc:neggran_pmc", "posdiv_pmc:negdiv_pmc"),
                    c("term", "estimate", "p.value")]


# WITH Affective Tone - interactions only
print("WITH Affective Tone - Interactions:")
s3_pool_results[s3_pool_results$term %in% 
                c("posgran_pmc:neggran_pmc", "posdiv_pmc:negdiv_pmc"),
                c("term", "estimate", "p.value")]
```


#plot
```{r}
nature_theme <- theme_classic() +
  theme(
    text = element_text(family = "serif"),
    axis.title = element_text(size = 12, family = "serif"),
    axis.text = element_text(size = 11, color = "black", family = "serif"),
    legend.title = element_text(size = 11, face = "bold", family = "serif"),
    legend.text = element_text(size = 10, family = "serif"),
    plot.title = element_text(size = 13, hjust = 0.5, family = "serif"),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "grey40"),
    plot.caption = element_text(size = 10, color = "grey50", hjust = 0),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.4),
    axis.ticks = element_line(color = "black", linewidth = 0.4),
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "plain"),   
    legend.key.width = unit(1.8, "cm")                     
  )

# Interaction plot
interact_plot(
  model_list_main_no_affective[[1]],
  pred = posdiv_pmc,
  modx = negdiv_pmc,
  interval = TRUE,
  int.width = 0.95,
  colors = c("#2166AC", "#4DAC26", "#D6604D"),
  x.label = "Positive diversity (person-mean-centered)",
  y.label = "Momentary Valence",
  legend.main = "Negative diversity",
  modx.labels = c("-1 SD (b = 2.09)",
                  "Mean (b = 2.59)",
                  "+1 SD (b = 3.08)"),
  plot.points = FALSE,
  vary.lty = FALSE, #Forces all lines to be solid
  pred.values = seq(-3, 3, by =0.1), #uniform x axis spacing
  pred.point.size = 0,
  robust = FALSE
) +
  nature_theme +
  labs(
    title = "Positive × Negative diversity Interaction"
  ) +
  annotate("text", x = Inf, y = Inf,
           label = "β = 0.47, p < .001",
           hjust = 1.1, vjust = 1.8,
           size = 4, fontface = "italic",
           color = "grey30") +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  guides(color = guide_legend(
    override.aes = list(linewidth = 1.2),
    keywidth = unit(1.5, "cm")
  ),
  linetype = "none" 
  )

ggsave("emodiversity_interaction.png",
       width = 6.5, height = 4.0, dpi = 300, bg = "white")

```


#PERSON LEVEL RESULTS
```{r}
#Load External Files - diversity and WhoQOL and Join
#This file contains combined person level posemo, negemo, diversity, MH scores
d_person_diversity <- read.csv("Person level diversity.csv", fileEncoding='UTF-8-BOM', sep = ';') %>%
  select(alias, posemo_person_avg, negemo_person_avg, positive_diversity_lem, negative_diversity_lem, Mental_mean)

print(names(d_person_diversity))
```


```{r}
# Clean alias function
clean_alias <- function(x) {
  x %>%
    str_to_lower() %>%
    str_replace_all("[^a-z0-9]", "")
}

# Step 2: Join using cleaned alias
person_level_list <- lapply(person_level_list, function(df) {
    df %>%
      mutate(alias_clean = clean_alias(alias)) %>%
      left_join(
        d_person_diversity %>% 
          mutate(alias_clean = clean_alias(alias)) %>%
          select(-alias),  # drop original alias from d_person_diversity
        by = "alias_clean"
      ) %>%
      select(-alias_clean)  # drop temp key after joining
  })
```


```{r}
# Step 3: # Convert list of 20 person-level datasets back to mids object
person_mids <- bind_rows(
  # Add the original data as .imp = 0 (required by as.mids)
  person_level_list[[1]] %>% mutate(.imp = 0, .id = row_number()),
  # Add all 20 imputed datasets with their index
  lapply(seq_along(person_level_list), function(i) {
    person_level_list[[i]] %>% mutate(.imp = i, .id = row_number())
  }) %>% bind_rows()
) %>%
as.mids()
```


```{r}
# Step 4: Standardize within each .imp at person level
person_long <- mice::complete(person_mids, action = "long", include = TRUE)

person_long_std <- person_long %>%
  group_by(.imp) %>%
  group_modify(~ datawizard::standardize(.x, select = c(
                                                "posemo_person_avg",
                                                "negemo_person_avg",
                                                "positive_diversity_lem",
                                                "negative_diversity_lem",
                                                "posgran_person_avg_lemm",
                                                "neggran_person_avg_lemm",
                                                "Mental_mean"))) %>%
  ungroup()

# Step 5: Convert back to mids
person_mids_std <- mice::as.mids(person_long_std)
```


#Mental Well Being - Pooling Results
#Person level - With Affective Tone
```{r}
mr_wb_s3 <- with(person_mids_std, lm(Mental_mean ~ posemo_person_avg * negemo_person_avg +
                             positive_diversity_lem * negative_diversity_lem+
                             posgran_person_avg_lemm * neggran_person_avg_lemm))

s3_person_pool <- mice::pool(mr_wb_s3)
s3_person_results <- summary(s3_person_pool, conf.int = TRUE)
s3_person_results
```


#person level - Without Affective Tone
```{r}
mr_wb_s4 <- with(person_mids_std, lm(Mental_mean ~ positive_diversity_lem * negative_diversity_lem +
                                                   posgran_person_avg_lemm * neggran_person_avg_lemm))

s4_person_pool <- mice::pool(mr_wb_s4)
s4_person_results <- summary(s4_person_pool, conf.int = TRUE)
s4_person_results
```



```{r}
# Build model list for person-level visualization
model_list_person <- map(
  seq_len(person_mids_std$m),
  \(i) {
    dat_i <- mice::complete(person_mids_std, action = i)
    lm(Mental_mean ~ positive_diversity_lem * negative_diversity_lem +
                     posgran_person_avg_lemm * neggran_person_avg_lemm,
       data = dat_i)
  }
)

# Marginal effect of positive granularity
model_list_person |>
  map(\(x) estimate_relation(x, by = "posgran_person_avg_lemm")) |>
  pool_predictions() |>
  plot(show_data = TRUE) +
  labs(title = "Positive Granularity and Mental Wellbeing",
       x = "Positive Granularity (person avg)",
       y = "Mental Wellbeing")

# Marginal effect of negative granularity
model_list_person |>
  map(\(x) estimate_relation(x, by = "neggran_person_avg_lemm")) |>
  pool_predictions() |>
  plot(show_data = TRUE) +
  labs(title = "Negative Granularity and Mental Wellbeing",
       x = "Negative Granularity (person avg)",
       y = "Mental Wellbeing")
```


```{r}
dat_viz <- mice::complete(person_mids_std, action = 1)

model_viz_person <- lm(Mental_mean ~ positive_diversity_lem * negative_diversity_lem +
                                     posgran_person_avg_lemm * neggran_person_avg_lemm,
                       data = dat_viz)

# Positive granularity
p1 <- effect_plot(model_viz_person,
                  pred = posgran_person_avg_lemm,
                  interval = TRUE,
                  partial.residuals = TRUE,
                  x.label = "Positive Granularity (person avg)",
                  y.label = "Mental Wellbeing",
                  colors = "royalblue1",
                  line.colors = "black",
                  point.size = 1.5,
                  point.alpha = 0.3,
                  jitter = 0.01)
p1 + theme(axis.title = element_text(size = 16))

# Negative granularity
p2 <- effect_plot(model_viz_person,
                  pred = neggran_person_avg_lemm,
                  interval = TRUE,
                  partial.residuals = TRUE,
                  x.label = "Negative Granularity (person avg)",
                  y.label = "Mental Wellbeing",
                  colors = "royalblue1",
                  line.colors = "black",
                  point.size = 1.5,
                  point.alpha = 0.3,
                  jitter = 0.01)
p2 + theme(axis.title = element_text(size = 16))
```


#Interactions without controlling for affective tone
```{r}
# Diversity interaction - person level (same style as text level plot)
interact_plot(
  model_viz_person,
  pred = positive_diversity_lem,
  modx = negative_diversity_lem,
  interval = TRUE,
  x.label = "Positive Diversity (person avg)",
  y.label = "Mental Wellbeing",
  legend.main = "Negative Diversity"
)

# Granularity interaction - person level
interact_plot(
  model_viz_person,
  pred = posgran_person_avg_lemm,
  modx = neggran_person_avg_lemm,
  interval = TRUE,
  x.label = "Positive Granularity (person avg)",
  y.label = "Mental Wellbeing",
  legend.main = "Negative Granularity"
)
```
